### 2023-03-03 chapters/abstract.tex
addressing Kuldeep's comments

	In interpretable machine learning, rule-based classifiers are particularly effective in representing the decision boundary using a set of rules. The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice. As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets. To this end, we propose an incremental learning framework, called {\imli}, which applies an iterative solving of  maximum satisfiability (MaxSAT) queries in mini-batch learning and enables classification on million-size datasets. Although being interpretable, rule-based classifiers often suffer from limited expressiveness, for example,  classifiers based on propositional logic. To learn more expressible yet interpretable classification rules, we propose a relaxation of classifiers based on logical formulas. For learning relaxed rule-based classifiers, we propose  an efficient learning framework, called {\crr},  building on incremental learning and mixed integer linear programming (MILP). {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.

	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers. In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features. Building on  stochastic satisfiability (SSAT), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features. For a more accurate fairness verification, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate  verification of fairness.

	Fairness metrics globally quantify bias, but do not detect or interpret its sources. To interpret group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier. FIF  interprets fairness by revealing potential individual or intersectional features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.

>>>>

	In interpretable machine learning, rule-based classifiers are particularly effective in representing the decision boundary using a set of rules. The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice. As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets. To this end, we discuss an incremental learning framework, called {\imli}, which applies an iterative solving of  maximum satisfiability (MaxSAT) queries in mini-batch learning and enables classification on million-size datasets. Although being interpretable, rule-based classifiers often suffer from limited expressiveness, for example,  classifiers based on propositional logic. To learn more expressible yet interpretable classification rules, we discuss a relaxation of classifiers based on logical formulas. For learning relaxed rule-based classifiers, we discuss an efficient learning framework, called {\crr},  building on incremental learning and mixed integer linear programming (MILP). {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.

	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers. In the presence of multiple fairness metrics for quantifying bias, we discuss a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features. Building on  stochastic satisfiability (SSAT), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features. For a more accurate fairness verification, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate  verification of fairness.

	Fairness metrics globally quantify bias, but do not detect or interpret its sources. To interpret group-based fairness metrics, we discuss fairness influence function (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier. FIF  interprets fairness by revealing potential individual or intersectional features attributing highly to the bias. Building on global sensitivity analysis, we discuss an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.

### 2023-03-03 chapters/background.tex
addressing Kuldeep's comments

\subsection{Propositional Satisfiability}

@@

A hard-OR or a soft-AND clause is a tuple $ (\propclause,\card) $ with an extra parameter $ \card $, where $ \card $ is the threshold on the literals in $ C $. Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $. Motivated by Boolean cardinality constraints, we propose relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$. We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{V \in \propclause_i} \mathds{1}[\sigma^* \models V] \geq \card_l$. Informally,  $\sigma^*$ satisfies a clause $(\propclause_i, \card_l)$ if  at least $\card_l$ literals in $\propclause_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(\propformula,\card_c, \card_l)$ if at least $\card_c$  clauses out of all $\{(\propclause_i,\card_l)\}_{i=1}^k$ clauses are true.

@@

\subsection{MaxSAT}

@@

\section{Interpretable Machine Learning}
\label{chapter_interpretability_preliminaries}
We consider a standard binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.  Here, the feature valuation vector $ \mathbf{x} = [x_1, \dots, x_{m}] \in \{0, 1\}^m $ is a vector of $ m $ Boolean features with $ y \in \{0,1\} $ being the binary class label. Thus, $ (\mathbf{x}, y) $ is called a positive sample if $ y = 1 $, and a negative sample otherwise.  We use $ (\nonsensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, y) $. Hence, each feature $ X_j \in \nonsensitive $ is sampled from a Bernoulli distribution $ \mathcal{X}_j $, and $ \mathcal{D} = \prod_{j=1}^m \mathcal{X}_j  $ is the product distribution of all features.

>>>>

\subsection{Propositional Satisfiability (SAT)}

@@

A hard-OR or a soft-AND clause is a tuple $ (\propclause,\card) $ with an extra parameter $ \card $, where $ \card $ is the threshold on the literals in $ C $. Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $. Motivated by Boolean cardinality constraints, we introduce relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$. We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{V \in \propclause_i} \mathds{1}[\sigma^* \models V] \geq \card_l$. Informally,  $\sigma^*$ satisfies a clause $(\propclause_i, \card_l)$ if  at least $\card_l$ literals in $\propclause_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(\propformula,\card_c, \card_l)$ if at least $\card_c$  clauses out of all $\{(\propclause_i,\card_l)\}_{i=1}^k$ clauses are true.

@@

\subsection{Maximum Satisfiability (MaxSAT)}

@@

\section{Interpretable Machine Learning}
\label{chapter_interpretability_preliminaries}
We consider a  binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.  Here, the feature valuation vector $ \mathbf{x} = [x_1, \dots, x_{m}] \in \{0, 1\}^m $ is a vector of $ m $ Boolean features with $ y \in \{0,1\} $ being the binary class label. Thus, $ (\mathbf{x}, y) $ is called a positive sample if $ y = 1 $, and a negative sample otherwise.  We use $ (\nonsensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, y) $. Hence, each feature $ X_j \in \nonsensitive $ is sampled from a Bernoulli distribution $ \mathcal{X}_j $, and $ \mathcal{D} = \prod_{j=1}^m \mathcal{X}_j  $ is the product distribution of all features.

### 2023-03-03 chapters/conclusion.tex
addressing Kuldeep's comments

Our work on fairness in machine learning introduces two novel probabilistic fairness verifiers, {\justicia} and {\fvgm}, which exhibit superior performance in accuracy and scalability compared to state-of-the-art verifiers. {\justicia} is a stochastic-SAT-based verifier that enables scalable verification of fairness for compound sensitive groups of Boolean classifiers, such as decision trees, which was previously infeasible with existing methods. On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers. Additionally, we propose a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features. Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.

>>>>

Our work on fairness in machine learning introduces two novel probabilistic fairness verifiers, {\justicia} and {\fvgm}, which exhibit superior performance in accuracy and scalability compared to state-of-the-art verifiers. {\justicia} is a stochastic-SAT-based verifier that enables scalable verification of fairness for compound sensitive groups of Boolean classifiers, such as decision trees, which was previously infeasible with existing methods. On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers. Additionally, we discuss a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features. Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.

### 2023-03-03 chapters/fairness/fif/appendix.tex
addressing Kuldeep's comments

\subsection{Accurate Approximation of Equalized Odds and Predictive Parity using FIFs.}
The accuracy of approximating equalized odds and predictive parity for {\fairXplainer} and SHAP is compared in Table~\ref{fairness_fairXplainer_tab:accuracy_eo_pp}. {\fairXplainer} shows lower estimation error compared to SHAP, especially when $\lambda=2$. SHAP is unable to explain predictive parity as predictive parity relies on the ground label $Y$, which is not available for randomly generated data points by SHAP for estimating local explanations.

@@

\subsection{Execution Time for FIF Estimation of Equalized Odds and Predictive Parity}
In Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}, we demonstrate the execution time of different methods in estimating FIFs of equalized odds and predictive parity in cactus plots. {\fairXplainer} with $ \lambda = 1 $ is more efficient than $ \lambda = 2 $ and solves all $ 480 $ fairness instances with at least one order of magnitude less execution time. Compared to {\fairXplainer} ($ \lambda = 1 $), SHAP demonstrates less computational efficiency.

>>>>

\subsection{Accuracy: Equalized Odds \& Predictive Parity via FIFs.}
The accuracy of approximating equalized odds and predictive parity for {\fairXplainer} and SHAP is compared in Table~\ref{fairness_fairXplainer_tab:accuracy_eo_pp}. {\fairXplainer} shows lower estimation error compared to SHAP, especially when $\lambda=2$. SHAP is unable to explain predictive parity as predictive parity relies on the ground label $Y$, which is not available for randomly generated data points by SHAP for estimating local explanations.

@@

\subsection{Execution Time: Equalized Odds \& Predictive Parity via FIFs}
In Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}, we demonstrate the execution time of different methods in estimating FIFs of equalized odds and predictive parity in cactus plots. {\fairXplainer} with $ \lambda = 1 $ is more efficient than $ \lambda = 2 $ and solves all $ 480 $ fairness instances with at least one order of magnitude less execution time. Compared to {\fairXplainer} ($ \lambda = 1 $), SHAP demonstrates less computational efficiency.

### 2023-03-03 chapters/fairness/fvgm/framework2.tex
addressing Kuldeep's comments

	\subsection{Fairness Verification with Computed Probability of Positive Prediction} 
	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair for $\epsilon \in [0,1]$ is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$. We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum probability of positive prediction $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum probability of positive prediction $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.

>>>>

	\subsection{Fairness Verification using Probability of Positive Prediction} 
	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair for $\epsilon \in [0,1]$ is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$. We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum probability of positive prediction $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum probability of positive prediction $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.

### 2023-03-03 chapters/fairness/justicia/discussion.tex
addressing kuldeep's comments
shortening
addressing Kuldeep's comments

\section{Chapter Summary}
Formal verification of different fairness metrics of machine learning for different datasets is an important question. Existing fairness verifiers, however, are not scalable, accurate, and extendable to non-Boolean sensitive features. We propose a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups.
Experimental evaluations demonstrate that {\justicia} achieves \textit{higher accuracy} and \textit{scalability} in comparison to the state-of-the-art verifiers, FairSquare and VeriFair, while yielding \textit{higher robustness} than the sample-based tools, such as AIF360.

Our work opens up several new directions of research. One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs. Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
%Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

>>>>

\section{Chapter Summary}
Formal verification of different fairness metrics of machine learning for different datasets is an important question. Existing fairness verifiers, however, are not scalable, accurate, and extendable to non-Boolean sensitive features. We discuss a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups. Experimental evaluations demonstrate that {\justicia} achieves \textit{higher accuracy} and \textit{scalability} in comparison to the state-of-the-art verifiers, FairSquare and VeriFair, while yielding \textit{higher robustness} than the sample-based tools, such as AIF360.  A limitation of {\justicia} is the lack of consideration of feature correlations in the SSAT-based verifier, which we alleviate in the next chapter.

%Our work opens up several new directions of research. One direction is to develop SSAT models and verifiers for popular classifiers like deep networks and SVMs. Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design. 
%Finally, an interesting direction is to enable {\justicia} to consider correlated feature distribution, which we pursue in the next chapter.
%Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2023-03-03 chapters/fairness/justicia/framework.tex
addressing Kuldeep's comments

\subsection{Inference Approach using ER-SSAT Encoding}
\label{fairness_justicia_sec:learn_ssat}
In most practical problems, there can be exponentially many compound sensitive groups due to the combination of categorical sensitive features.  As a result, the enumeration approach may suffer from scalability issues due to the exponential number of calls to the SSAT solver. To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier. As discussed above, these two probabilities are sufficient to measure different group and causal fairness metrics.

>>>>

\subsection{Inference Approach using ER-SSAT Encoding}
\label{fairness_justicia_sec:learn_ssat}
In most practical problems, there can be exponentially many compound sensitive groups due to the combination of categorical sensitive features.  As a result, the enumeration approach may suffer from scalability issues due to the exponential number of calls to the SSAT solver. To this end, we discuss an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier. As discussed above, these two probabilities are sufficient to measure different group and causal fairness metrics.

### 2023-03-03 chapters/fairness/justicia/introduction.tex
outline in imli
addressing Kuldeep's comments

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
In this chapter, we study formal fairness verification problem in machine learning, where we verify the bias of a classifier given the probability distribution of features. To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

@@

\paragraph{Contribution.} We propose to model verifying different fairness metrics as a SSAT problem.  We primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   Our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

@@

Our contributions are summarised below:
\begin{itemize}
	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
\end{itemize}

@@

\begin{figure}[!t]
	\centering
	\subfloat[Dependency among features and prediction]{\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.6cm]
				% Define nodes
				\node[latent,scale=1] (a1) {$\textrm{age}$} ; %
				\node[obs, scale=1, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
				\node[obs, scale=1, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
				\node[obs, scale=1, below=of h, xshift=.8cm] (p) {$\widehat{Y}$}; %			
				%%add edge
				\edge[] {a1} {h,i} ;
				\edge[] {h,i} {p} ;
			\end{tikzpicture}
		}\label{fairness_justicia_fig:fair_example_dependency}}\hfill
	\subfloat[Age-dependent distributions of non-sensitive features]{
		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
		\label{fairness_justicia_fig:fair_example_distribution}}

	\subfloat[Trained decision tree]{
		\scalebox{0.55}{	
			\begin{tikzpicture}[x=1cm,y=1.8cm]
				\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
				\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
				\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
				\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
				%
				\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
				\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
				\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
				\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
				\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
				\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
		\end{tikzpicture}}
		\label{fairness_justicia_fig:fair_example_dt}}\hfill
	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
		\scalebox{1}{
			\begin{tikzpicture}
				\begin{axis}[
					enlarge x limits=0.2,
					symbolic x coords={$ \text{age} \ge 40 $, $ \text{age} < 40 $},
					xtick={$ \text{age} \ge 40 $, $ \text{age} < 40 $},  
					%			xticklabel style={text height=2ex}, 
					ylabel={$ \Pr[\hat{Y} =  1] $},
					xlabel={Sensitive groups},
					ymin=0,
					width= 0.3\textwidth,]

>>>>

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
We discuss formal fairness verification problem in machine learning where we verify the bias of a classifier given the probability distribution of features. To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

@@

\paragraph{Contribution.} We discuss model verifying different fairness metrics as a SSAT problem.  We primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   Our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

@@

Our contributions are summarised below:
\begin{itemize}
	\item We discuss a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
	\item Unlike earlier probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
\end{itemize}

@@

\begin{figure}[!t]
	\centering
	\subfloat[Dependency among features and prediction]{\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.6cm]
				% Define nodes
				\node[latent,scale=1] (a1) {$\textrm{age}$} ; %
				\node[obs, scale=1, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
				\node[obs, scale=1, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
				\node[obs, scale=1, below=of h, xshift=.8cm] (p) {$\widehat{Y}$}; %			
				%%add edge
				\edge[] {a1} {h,i} ;
				\edge[] {h,i} {p} ;
			\end{tikzpicture}
		}\label{fairness_justicia_fig:fair_example_dependency}}\hfil
	\subfloat[Age-dependent distributions of non-sensitive features]{
		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
		\label{fairness_justicia_fig:fair_example_distribution}}

	\subfloat[Trained decision tree]{
		\scalebox{0.55}{	
			\begin{tikzpicture}[x=1cm,y=1.8cm]
				\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
				\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
				\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
				\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
				%
				\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
				\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
				\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
				\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
				\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
				\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
		\end{tikzpicture}}
		\label{fairness_justicia_fig:fair_example_dt}}\hfil
	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
		\scalebox{1}{
			\begin{tikzpicture}
				\begin{axis}[
					enlarge x limits=0.2,
					symbolic x coords={$ \text{age} \ge 40 $, $ \text{age} < 40 $},
					xtick={$ \text{age} \ge 40 $, $ \text{age} < 40 $},  
					%			xticklabel style={text height=2ex}, 
					ylabel={$ \Pr[\hat{Y} =  1] $},
					xlabel={Sensitive groups},
					ymin=0,
					width= 0.3\textwidth,]

### 2023-03-03 chapters/interpretability/imli/appendix.tex
addressing Kuldeep's comments

\section{Experimental Comparison between Incremental vs.\ Non-incremental  Encoding}
We compare the performance of two encoding techniques, na\"ive non-incremental encoding and efficient incremental encoding in {\imli}, for learning interpretable CNF classification rules in datasets of various sizes. We demonstrate the scalability results using a cactus plot in Figure~\ref{fig:interpretability_imli_additional_fig:scalability_non_incremental_vs_incremental}. Each point $ (x, y) $ represents the ability to solve $ x $ classification instances within $ y $ seconds. We test both incremental and non-incremental encoding using $ 360 $ instances for each dataset with varying hyper-parameters. As the dataset size increases, the non-incremental encoding times out, failing to solve all $ 360 $ instances within $ 1000 $ seconds in the Credit and Adult datasets. In contrast, the incremental encoding in {\imli} demonstrates higher scalability, solving all $ 360 $ instances in less than $ 10 $ seconds in the Titanic dataset, and efficiently solving all $ 360 $ instances within the timeout in the Credit and Adult datasets. Hence, the incremental encoding is superior in terms of scalability compared to the non-incremental encoding.

>>>>

\section{Performance Comparison: Incremental vs.\ Non-incremental  Encoding}
We compare the performance of two encoding techniques, na\"ive non-incremental encoding and efficient incremental encoding in {\imli}, for learning interpretable CNF classification rules in datasets of various sizes. We demonstrate the scalability results using a cactus plot in Figure~\ref{fig:interpretability_imli_additional_fig:scalability_non_incremental_vs_incremental}. Each point $ (x, y) $ represents the ability to solve $ x $ classification instances within $ y $ seconds. We test both incremental and non-incremental encoding using $ 360 $ instances for each dataset with varying hyper-parameters. As the dataset size increases, the non-incremental encoding times out, failing to solve all $ 360 $ instances within $ 1000 $ seconds in the Credit and Adult datasets. In contrast, the incremental encoding in {\imli} demonstrates higher scalability, solving all $ 360 $ instances in less than $ 10 $ seconds in the Titanic dataset, and efficiently solving all $ 360 $ instances within the timeout in the Credit and Adult datasets. Hence, the incremental encoding is superior in terms of scalability compared to the non-incremental encoding.

### 2023-03-03 chapters/interpretability/imli/conclusion.tex
shortening
addressing Kuldeep's comments

\section{Conclusion}
\label{interpretability_imli_sec:conclusion}
Interpretable machine learning is gaining more focus with applications in many safety-critical domains. Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.  In this paper, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas. {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.  In our empirical evaluation, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. In particular, {\imli} demonstrates competitive prediction accuracy and rule-size compared to existing interpretable rule-based classifiers. In addition, {\imli} achieves impressive scalability than both interpretable and non-interpretable classifiers by learning interpretable rules on million-size datasets with higher accuracy.  Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework.

>>>>

\section{Chapter Summary}
\label{interpretability_imli_sec:conclusion}
Interpretable machine learning is gaining more focus with applications in many safety-critical domains. Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.  In this chapter, we discuss a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas. {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.  In our empirical evaluation, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. In particular, {\imli} demonstrates competitive prediction accuracy and rule-size compared to existing interpretable rule-based classifiers. In addition, {\imli} achieves impressive scalability than both interpretable and non-interpretable classifiers by learning interpretable rules on million-size datasets with higher accuracy.  Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework. In the next chapter, we leverage the incremental learning in {\imli} to learn more expressible yet interpretable classification rules.
